---
title: "Data Visualization"
author: "Cody, Max, Priya, and Mohammed"
output:
  html_document:
    fig_caption: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F,
                      message = F)
```

Introduction

Baseball, sometimes referred to as the Great American Pastime, is one of the most popular sports in America. There is an immense amount of baseball data freely available for analysis and is effectively a statistician’s playground. 

We chose the topic of baseball for our final project for a few reasons. One of our group members is involved in fantasy baseball, and we thought that exploring baseball data in-depth would probably help him in future fantasy bracket player selections. We also hoped that this analysis, or a similar analysis, would be potentially applicable to other sports including but not limited to basketball and American football. Lastly we wanted to get insight into a core element of American pop culture, and look into some interesting questions such as steroid use. We had a lot of questions that we thought of, and have listed these below:

Are there any obvious trends between pitching volume and effort in a given year and injuries in the next year?
*How severe are different kinds of injuries to a baseball player’s career?
*Are pitchers who throw different kinds of pitches more susceptible to getting injured? Do certain kinds of pitches lead to different kinds of injuries?
*How do different kinds of injuries affect hitting power?
*How does likelihood of future injury change with different kinds of injuries?
*How did a ball manufacturing change affect number of home runs?
*How may a robot umpire affect the strike zone?

Our group members are:
*Cody Zupnick - data collection, plotting, and report writing
*Mohammad Radiyat - plotting and report writing
*Priya Medberry - 
*Max Mattioli - data quality analysis and report writing

Description of data: 

The data was collected from a few different sources. Pitcher data was collected from the following [link.](www.fangraphs.com%2Fleaders.aspx%3Fpos%3Dall%26stats%3Dpit%26lg%3Dall%26qual%3Dy%26type%3Dc%252C3%252C31%252C125%252C126%252C127%252C128%252C130%252C131%252C132%252C133%252C135%252C139%252C13%26season%3D2017%26month%3D0%26season1%3D2017%26ind%3D0%26team%3D%26rost%3D%26age%3D%26filter%3D%26players%3D&h=ATM_yMkJmnVUTRMPwJ4QLEVH8tYNAjxgBIYrmH6RqQIzYc_-MeKG_silRZlJQeofqEGo32CC01V4zqagZT4DPi5eIgG67J__srp7yg) The batter data was collected from [here.](www.fangraphs.com%2Fleaders.aspx%3Fpos%3Dall%26stats%3Dbat%26lg%3Dall%26qual%3Dy%26type%3D8%26season%3D2017%26month%3D0%26season1%3D2017%26ind%3D0%26team%3D0%26rost%3D0%26age%3D0%26filter%3D%26players%3D0&h=ATM_yMkJmnVUTRMPwJ4QLEVH8tYNAjxgBIYrmH6RqQIzYc_-MeKG_silRZlJQeofqEGo32CC01V4zqagZT4DPi5eIgG67J__srp7yg)

The injury data was the most difficult to collect. Injury data was obtained via scraping, and the scripts to do so can be found at this Github [link.](https://github.com/Codyz/edavFinalProject/tree/master/scrape)



```{r, cache=T}
# Install all necessary packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(zoo)
```

**Part 1: Pitching Volume, Effort, and Injuries**

Here we're going to take a look to see if any obvious trends can be seen between pitching volume in year N and injuries in year N+1. Our metrics of volume are seasonal pitch counts, pitches per game, and innings pitched per season. Our basic metric of effort is fastball velocity. All of these are imperfect metrics; a stronger pitcher can throw harder with less effort than a weaker one, for instance, and not all pitches are equally strenuous. But this is a good starting points. Another consideration we're not immediately considering is age. We'll take a look at this later, because we expect it to be very important, since older players get injured more frequently than younger ones. 

```{r}
pitches_2015 <- read_csv('./data/Pitchers2015.csv')
injuries_2016 <- read_csv('./data/injuries_2016.csv')

pitches_2016 <- read_csv('./data/Pitchers2016.csv')
injuries_2017 <- read_csv('./data/injuries_2017.csv')

# join each year before concatting them
joined1 <- merge(injuries_2016, pitches_2015, by.x = "name", by.y = "name", all.y = T)
joined1$year <- '2016'

joined2 <- merge(injuries_2017, pitches_2016, by.x = "name", by.y = "name", all.y = T)
joined2$year <- '2017'

joined <- rbind(joined1, joined2)

joined_pitchers <- filter(joined, pos == "RP" | pos == "SP")

# if a player has multiple injuries, encoded as Injury1/Injury2/... First is most serious, so just consider that
extract_injury <- function(val) { strsplit(val, "/")[[1]][[1]] }

# clean up the data
# @param df pitch dataframe joined with injury data
clean <- function(df) {
  percent_cols <- colnames(df)[8:16]
  f <- function(val) { sub("%", "", val) }
  df[percent_cols] <- sapply(df[percent_cols], f)
  df[percent_cols] <- sapply(df[percent_cols], as.numeric)
  # if a pitcher never throws a pitch
  df[percent_cols][is.na(df[percent_cols])] <- 0
  df$type[is.na(df$type)] <- "None"
  df$days[is.na(df$days)] <- 0

  df$type <- sapply(df$type, extract_injury)
  
  df
}

# lastly, let's only consider injury types suffered by at least 10 players
joined <- clean(joined)
joined_pitchers <- clean(joined_pitchers)
joined_pitchers <- mutate(group_by(joined_pitchers, type), cnt = n())
joined_pitchers <- filter(joined_pitchers, cnt >= 10)
```

```{r}
ggplot(joined_pitchers, aes(x=type, y=Pitches)) + geom_jitter() + coord_flip()
ggplot(joined_pitchers, aes(x=type, y=vFA)) + geom_jitter() + coord_flip()
```

So this doesn't really look like anything. A very common thought is that high pitch total pitch volumes lead to injuries, but we can't see it over the course of one season, or it's being
swamped by other factors. Again, nothing obvious jumps out. 

##Injury Type

Let's take a quick look at how severe differnt injuries are, across all players in season.

```{r}
ggplot(filter(joined, days < 180), aes(x=type, y=days)) + geom_jitter() + coord_flip()
```
Shoulder, arm, and elbow injuries look bad. More important than the mean values, we can see that pitching injuries are the most common, The densest clusters are for shoulder, elbow, arm, and back.


## Types of Pitches And Injuries
Are pitchers who throw certain types of pitches more susceptible to getting injured? And do certain kinds of pitches lead to different types of injuries? This has been a holy grail of baseball research, and it's still hotly debated, so we won't settle anything without modeling, but maybe we can see something.  

```{r}
percent_cols <- colnames(joined)[8:16]
# tidy the data so we can facet on pitch types
g <- gather(filter(joined_pitchers), pt, pt_pct, percent_cols)
levels(g$pt) <- percent_cols
g <- arrange(g, pt)
ggplot(g, aes(x = type, y = pt_pct, fill = pt, order = pt)) + geom_bar(stat="summary", position = "dodge") + theme(axis.text.x = element_text(angle=30))

percent_cols_no_FA <- colnames(joined)[9:16]
# tidy the data so we can facet on pitch types
g <- gather(filter(joined_pitchers), pt, pt_pct, percent_cols_no_FA)
levels(g$pt) <- percent_cols_no_FA
g <- arrange(g, pt)
ggplot(g, aes(x = type, y = pt_pct, fill = pt, order = pt)) + geom_bar(stat="summary", position = "dodge") + theme(axis.text.x = element_text(angle=30))
```
\newpage


**Part2: How Do Injuries Effect Power?**

Presumably some kinds of injuries hurt hitters more than others. We're going to take a look at how different kinds of injuries effect power performance. Because we're interested in the long
term effects of injuries, we're going to be looking at a one year lag, *ignoring* the season in which the player was injured. We're going to look at season before and seasons after the injury.

```{r}
# Prepare the data
injuries2015 <- read_csv('./data/injuries_2015.csv')
injuries2016 <- read_csv('./data/injuries_2016.csv')

df2014 <- read_csv('./data/batters2014.csv')
df2015 <- read_csv('./data/batters2015.csv')
df2016 <- read_csv('./data/batters2016.csv')
df2017 <- read_csv('./data/batters2017.csv')

df2014$year <- 2014
df2015$year <- 2015
df2016$year <- 2016
df2017$year <- 2017

# tidy the data for a three year period
process <- function(data, injuries) {
  # we only want players who played in each season, and batted at least 150 times in 2015/2017
  data <- mutate(group_by(data, name), cnt = n())
  first_yr <- min(data$year)
  third_yr <- max(data$year)
  scnd_year <- (first_yr + third_yr) / 2
  data <- filter(data, cnt == 3 & year != scnd_year)
  # tidy till before and after
  data <- mutate(data, occ = if_else(year == first_yr, "before", "after"))
  data <- data %>% select(name, occ, SLG)
  data <- spread(data, occ, SLG)
  # merge the injuries
  data <- merge(data, injuries, by.x = "name", by.y = "name", all.x = T)
  data$type[is.na(data$type)] <- "None"
  data$type <- sapply(data$type, extract_injury)
  data
}

bat1 <- rbind(df2015, df2016, df2017)
bat2 <- rbind(df2014, df2015, df2016)

bat1 <- process(bat1, injuries2016)
bat2 <- process(bat2, injuries2015)

bat <- rbind(bat1, bat2)
```

```{r}
ggplot(bat, aes(before, after)) + geom_point(aes(color = type))
```

It's hard to see much. Let's try a bar chart, plotting the ratios of before/after.
```{r}
bat <- mutate(bat, chng = before / after)
ggplot(bat) + geom_bar(aes(x = type, y = chng), stat="summary") + stat_summary(aes(x = type, y = chng), fun.y = "mean") + coord_flip()
```

Now it's possible to see some patterns emerge. Lower body injuries seem to have the largest effect on power; ACL tears, knee injuries, toe injuries, hip, and calf injuries have the largest decreases. This makes intuitive sense, since a lot of hitting is putting your lower half into the swing. But some of this is non-obvious; we wouldn't have expected a toe injury to have a larger effect on how hard a player
hits the ball than a wrist injury. Of course this is only four seasons of data, and we'd want to revisit this on larger datasets. But the lower body/upper body difference seems strong.  


**Part 3: How Well Do Injuries Predict Future Injuries?**

```{r}
# process injury data
i1 <- read_csv('./data/injuries_2015.csv')
i2 <- read_csv('./data/injuries_2016.csv')
i3 <- read_csv('./data/injuries_2017.csv')

i1$year = 2015
i2$year = 2016
i3$year = 2017

iAll <- rbind(i1, i2, i3)
iAll <- arrange(iAll, name, year)
iAll <- mutate(group_by(iAll, name), cnt = n())
# data <- mutate(data, occ = if_else(year == first_yr, "before", "after"))
iAll <- iAll %>% mutate(was_followed = if_else(cnt > 1 & year != 2017, 1, 0))
# remove 2017
iAll <- filter(iAll, year != 2017)
iAll$type <- sapply(iAll$type, extract_injury)
iAll <- iAll %>% group_by(type, was_followed) %>%
          summarise(num_i = n()) %>%
          mutate(prop = num_i / sum(num_i))
```

```{r}
# 2-D Mosaic Plot/Stacked bar chart
iAll <- iAll %>% mutate(followed = if_else(was_followed == 0, "No", "Yes"))
ggplot(iAll, aes(x = type, y = prop, fill = followed)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle=90))
```

The numbers aren't huge, since finding injury data back more than a few years is difficult, but it definitely looks like some injuries are more predictive of future injuries. With a larger dataset, we could
investigate if certain types of injuries predict *which* future injuries, but our slices would be too small for just three years. 


**Part 4: Did the Ball Change in the Middle of 2015?**
First, let's just establish that home runs *did* increase.

```{r}
## structure the data so that months are orderd included, in the date
hr <- read_csv('./data/homeRuns.csv')
# just make Mar/April April and Sep/Oct September
hr$Month[hr$Month == 'Mar/Apr'] <- 'Apr'
hr$Month[hr$Month == 'Sept/Oct'] <- 'Sep'
# zoo parser is much better
hr$Date <- as.yearmon(paste0(hr$Season, hr$Month), "%Y%b")
hr$Date <- as.Date(hr$Date)
```
```{r}
ggplot(hr, aes(Date, y = HR)) + geom_line() + scale_x_date(labels = as.Date(hr$Date))
```

**Part 5: How Might Robot Umpires Effect The Strike Zone?**

